Synthesizing Symmetric Lenses

### 2018-07-11 papers/SymmetricLenses/main.tex
Spacing fix applied
some edits to section 2

\subsection{Simple Symmetric Lenses}
A simple symmetric lens between formats described by types $X$ and $Y$ is
defined by four functions.

%\begin{enumerate}
%\item $\CreateR \OfType \Regex \rightarrow \RegexAlt$
%\item $\CreateL \OfType \RegexAlt \rightarrow \Regex$
%\item $\PutR \OfType \Regex \rightarrow \RegexAlt \rightarrow \RegexAlt$
%\item $\PutL \OfType \RegexAlt \rightarrow \Regex \rightarrow \Regex$
%\end{enumerate}

\begin{gather}
 \tag{1}
 \mbox{$\CreateR \OfType X \rightarrow Y$}\\
 \tag{2}
 \mbox{$\CreateL \OfType Y \rightarrow X$}\\
 \tag{3}
 \mbox{$\PutR \OfType X \rightarrow Y \rightarrow Y$}\\
 \tag{4}
 \mbox{$\PutL \OfType Y \rightarrow X \rightarrow X$}
\end{gather}

\noindent
that collectively satisfy four ``round-tripping'' laws: 
\begin{gather}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{gather}
\noindent
The two {\sc Create} functions propagate shared information and fill in default
values when introducing new data. (\EG, the ``unk'' salary entry is added
alongside the name to the management file when HR inserts a new employee). The
two {\sc Put} functions propagate edits from one format to the other by
combining the change from one file with the previous value from the other file.
The roundtripping laws guarantee that if two formats are already synchronized,
{\sc Put}s merely return the second argument, unchanged.

This definition differs from previous formulations of simple symmetric lenses,
as previous formulations required a \emph{complement}. Instead of comprising
four functions, existing formulations merely comprised two functions, which
combined complements with edited formats to create updated complements and
synchronized data. We provide a detailed comparison to classic symmetric lenses
in \S\ref{sec:relationship}.

@@

Generally, we do not expect programmers to define simple symmetric lenses by
writing the four functions by hand and showing they satisfy the roundtripping
laws. Instead, we provide simple symmetric lens combinators to build up the
requisite lenses from simpler ones, and focus our combinators on providing
lenses between string data formats.

@@

A challenge is that the simple symmetric lens combinators permit many
well-typed lenses
between a given pair of regular expressions. For example, 
Figure~\ref{fig:example_lens} gives one possible lens between regular
expressions  \lstinline{emp_salaries} and \lstinline{emp_insurance},
but 
%
\begin{lstlisting}
disconnect(emp_salaries,emp_insurance,rep(emp_salaries),rep(emp_insurance))
\end{lstlisting}
%
is another valid lens (where \lstinline{rep($\BRegex$)} denotes some string in the
language of $\BRegex$). Furthermore, adding the input-output example in
Figure~\ref{fig:minimized-representations} 
helps, but only a little: it may, in general, require {\em many} examples to rule
out all possible occurrences of nested \Disconnect lenses,
particularly in complex formats. 
%\ksf{I don't understand the following.  We shouldn't explain the issue in
%terms of \PutR and \PutL as we are focused on the combinators rather
%than the raw functions.}
%The default strings given to \Disconnect merely need to be the strings
%in the input-output examples, and the lens will still satisfy the
%specification.  While the disconnect lens is prohibited by two
%examples, or by a carefully chosen input-output example on \PutR
%or \PutL, this problem continues presenting itself for the sublenses,
%and is highlighted even more on larger formats.
While we could ask programmers to provide detailed and precise specifications
for their synthesis tasks, we would prefer an algorithm that can intelligently
choose a lens despite the underspecification. Instead of merely finding
\emph{any} satisfying lens, we wish to synthesize a satisfying lens that is
likely to please the user.

@@

Two classical symmetric lenses are equivalent if they output the same
formats given any sequence of edits. Formally, given a lens
$\Lens \in X \leftrightarrow Y$,
an \emph{edit} for $\Lens$ is a member of $X + Y$. Consider
the function $apply$, which, given a lens and an element of that lens's
complement, is a function from sequences of edits to sequences of edits. If
$apply(\Lens,c,es) = es'$, then that means, starting with complement $c$, after
edit $es_i$, the lens $\Lens$ synchronizes the other data to $es'_i$.
\[
  \inferrule
  {
  }
  {
    apply(\Lens,c,[]) = []
  }
\]
\[
  \inferrule
  {
    \Lens.putr(x,c) = (y,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InLOf{x})::es) = (\InROf{y})::es'
  }
\]
\[
  \inferrule
  {
    \Lens.putl(y,c) = (x,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InROf{y})::es) = (\InLOf{x})::es'
  }
\]
\noindent
Two lenses, $\Lens_1$ and $\Lens_2$, are equivalent if
$apply(\Lens_1,\Lens_1.init,es) = apply(\Lens_2,\Lens_2.init,es)$ for all edit
sequences $es$.

@@

\[
  \inferrule
  {
  }
  {
    apply(\Lens,xyo,[]) = []
  }
\]
\[
  \inferrule
  {
    \Lens.\CreateROf{x} = y\\
    apply(\Lens,\SomeOf{(x,y)},es) = es'
  }
  {
    apply(l,\None,\InLOf{x}::es) = \InROf{y}::es'
  }
\]
\[
  \inferrule
  {
    \Lens.\CreateLOf{y} = x\\
    apply(\Lens,\SomeOf{(x,y)},es) = es'
  }
  {
    apply(l,\None,\InROf{y}::es) = \InLOf{x}::es'
  }
\]
\[
  \inferrule
  {
    \Lens.\PutROf{x'}{y}  = y'\\
    apply(\Lens,\SomeOf{(x',y')},es) = es'
  }
  {
    apply(l,\SomeOf{(x,y)},\InLOf{x'}::es) = \InROf{y'}::es'
  }
\]
\[
  \inferrule
  {
    \Lens.\PutLOf{y'}{x}  = x'\\
    apply(\Lens,\SomeOf{(x',y')},es) = es'
  }
  {
    apply(l,\SomeOf{(x,y)},\InROf{y'}::es) = \InLOf{x'}::es'
  }
\]

>>>>

\subsection{Simple Symmetric Lenses}
A simple symmetric lens between formats described by types $X$ and $Y$ is
defined by four functions that collectively satisfy four ``round-tripping'' laws:

%\begin{enumerate}
%\item $\CreateR \OfType \Regex \rightarrow \RegexAlt$
%\item $\CreateL \OfType \RegexAlt \rightarrow \Regex$
%\item $\PutR \OfType \Regex \rightarrow \RegexAlt \rightarrow \RegexAlt$
%\item $\PutL \OfType \RegexAlt \rightarrow \Regex \rightarrow \Regex$
%\end{enumerate}
\vskip 1.5ex
\begin{minipage}[c]{.3\textwidth}
\begin{gather}
 \tag{1}
 \mbox{$\CreateR \OfType X \rightarrow Y$}\\
 \tag{2}
 \mbox{$\CreateL \OfType Y \rightarrow X$}\\
 \tag{3}
 \mbox{$\PutR \OfType X \rightarrow Y \rightarrow Y$}\\
 \tag{4}
 \mbox{$\PutL \OfType Y \rightarrow X \rightarrow X$}
\end{gather} 
\end{minipage} 
\begin{minipage}[c]{.6\textwidth}
\begin{align}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{align}
\end{minipage} 
\vskip 1ex
\noindent
The two \KW{create} functions propagate shared information and fill in default
values when introducing new data. (\EG, the ``unk'' salary entry is added
alongside the name to the management file when HR inserts a new employee). The
two \KW{put} functions propagate edits from one format to the other by
combining the change from one file with the previous value from the other file.
The round-tripping laws guarantee that if two formats are already synchronized,
\KW{put}s merely return the second argument, unchanged.

This definition differs from formulations of classic symmetric lenses,
which require a \emph{complement} but only two functions.  These two
functions combine complements with edited data to create updated
complements and synchronized data, one function for each direction.
We provide a detailed comparison to classic symmetric lenses
in \S\ref{sec:relationship}.

@@

Generally, we do not expect programmers to define simple symmetric lenses by
writing the four functions by hand and showing they satisfy the round-tripping
laws. Instead, we provide simple symmetric lens combinators to build up the
requisite lenses from simpler ones. We focus on combinators for defining
lenses between string data formats.

@@

A challenge is that the simple symmetric lens combinators permit many
well-typed lenses
between a given pair of regular expressions. For example, 
Figure~\ref{fig:example_lens} gives one possible lens between regular
expressions  \lstinline{emp_salaries} and \lstinline{emp_insurance},
but 
%
\begin{lstlisting}
disconnect(emp_salaries,emp_insurance,rep(emp_salaries),rep(emp_insurance))
\end{lstlisting}
%
is another valid lens (where \lstinline{rep($\BRegex$)} denotes some string in the
language of $\BRegex$). Furthermore, adding the input-output example in
Figure~\ref{fig:minimized-representations} 
helps, but only a little: It may, in general, require {\em many} examples to rule
out all possible occurrences of nested \Disconnect lenses,
particularly in complex formats. 
%\ksf{I don't understand the following.  We shouldn't explain the issue in
%terms of \PutR and \PutL as we are focused on the combinators rather
%than the raw functions.}
%The default strings given to \Disconnect merely need to be the strings
%in the input-output examples, and the lens will still satisfy the
%specification.  While the disconnect lens is prohibited by two
%examples, or by a carefully chosen input-output example on \PutR
%or \PutL, this problem continues presenting itself for the sublenses,
%and is highlighted even more on larger formats.
While we could ask programmers to provide detailed and precise specifications
for their synthesis tasks, we would prefer an algorithm that can intelligently
choose a lens despite the underspecification. Instead of merely finding
\emph{any} satisfying lens, we wish to synthesize a satisfying lens that is
likely to please the user.

@@

Two classical symmetric lenses are equivalent if they output the same
formats given any sequence of edits. Formally, given a lens
$\Lens \in X \leftrightarrow Y$,
an \emph{edit} for $\Lens$ is a member of $X + Y$. Consider
the function $apply$, which, given a lens and an element of that lens's
complement, is a function from sequences of edits to sequences of edits. If
$apply(\Lens,c,es) = es'$, then that means, starting with complement $c$, after
edit $es_i$, the lens $\Lens$ synchronizes the other data to $es'_i$.
\begin{centermath}
\begin{tabular}[b]{l@{\qquad \; \qquad}l}
$
  \inferrule
  {
  }
  {
    apply(\Lens,c,[]) = []
  }
$
&
$
  \inferrule
  {
    \Lens.putr(x,c) = (y,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InLOf{x})::es) = (\InROf{y})::es'
  }
$
\end{tabular}
\end{centermath}
\[
  \inferrule
  {
    \Lens.putl(y,c) = (x,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InROf{y})::es) = (\InLOf{x})::es'
  }
\]
\noindent
Two lenses, $\Lens_1$ and $\Lens_2$, are equivalent if
$apply(\Lens_1,\Lens_1.init,es) = apply(\Lens_2,\Lens_2.init,es)$ for all edit
sequences $es$.

@@

\begin{centermath}
\begin{tabular}[b]{l@{\qquad \; \qquad}l}
$
  \inferrule
  {
  }
  {
    apply(\Lens,xyo,[]) = []
  }
$
&
$
  \inferrule
  {
    \Lens.\CreateROf{x} = y\\
    apply(\Lens,\SomeOf{(x,y)},es) = es'
  }
  {
    apply(l,\None,\InLOf{x}::es) = \InROf{y}::es'
  }
$
\end{tabular}
\end{centermath}
\[
  \inferrule
  {
    \Lens.\CreateLOf{y} = x\\
    apply(\Lens,\SomeOf{(x,y)},es) = es'
  }
  {
    apply(l,\None,\InROf{y}::es) = \InLOf{x}::es'
  }
\]
\[
  \inferrule
  {
    \Lens.\PutROf{x'}{y}  = y'\\
    apply(\Lens,\SomeOf{(x',y')},es) = es'
  }
  {
    apply(l,\SomeOf{(x,y)},\InLOf{x'}::es) = \InROf{y'}::es'
  }
\]
\[
  \inferrule
  {
    \Lens.\PutLOf{y'}{x}  = x'\\
    apply(\Lens,\SomeOf{(x',y')},es) = es'
  }
  {
    apply(l,\SomeOf{(x,y)},\InROf{y'}::es) = \InLOf{x'}::es'
  }
\]
